---
title: "step6_timecourse"
date: "2025-07-21"
---

This script computes and visualises the time course of infants’ looking behaviour toward the deviant stimulus.
•	For each trial, gaze samples during the shapes_on_1 and shapes_on_2 events were coded as:
o	1 if gaze fell on the deviant stimulus location
o	0 if gaze fell on any standard stimulus location
(samples off-screen or unrelated were excluded).
•	For each participant, shape, event, and time sample, the proportion looking time at the deviant (PLTD) was computed as the mean of this binary gaze variable.
•	 These participant-level PLTDs were then averaged across participants for each shape and time sample.


# Setup

## Libraries

```{r}
# Import the necessary libraries
library(readxl)
library(tidyverse)
library(rstatix)
library(ggforce)
```

```{r}
exp <- 'E1'

#for plotting purpose
if (exp == "E1") {
  levels_labels <- c(
    "object",
    "equilateral_triangle",
    "isosceles_triangle",
    "square",
    "rhombus",
    "pentagon"
  )
} 
```

## Paths

```{r}
# Construct the paths
path_data_processed <- file.path(getwd(), paste0("data/processed_", exp))
path_wd <- file.path(getwd())
```

## Load data [eyetr w/ gaze locations]

```{r}
setwd(path_data_processed)
load(sprintf("data_output_step02_%s.RData", exp))
```

```{r}
data_output_step02 <- data_output_step02 %>%
  mutate(shape_w_duck_car = shape) %>%
  mutate(shape = ifelse(shape %in% c('duck','car'),'object',shape))

#for plotting purpose
desired_shape_order <- levels_labels

data_output_step02 <- data_output_step02 %>%
  mutate(shape = factor(shape, levels = desired_shape_order))
```
## Time course 

```{r}
aux <- data_output_step02 %>%
  filter(validtrial == 1) %>%
  filter(ppt_included == 1) %>%
  filter(event %in% c("shapes_on_1", "shapes_on_2")) %>%
  mutate(
    gaze_quadrant = as.numeric(str_extract(gaze, "\\d"))
  )

data_time_course <- aux %>%
  mutate(gaze_is_on_deviant = case_when(
    #gaze %in% c("offscreen", "nogaze") ~ NaN,
    gaze_quadrant == location_deviant ~ 1,
    gaze_quadrant %in% c(location_standard, location_standard_smaller, location_standard_larger) ~ 0  # standard
    )) %>%
  mutate(gaze_stim = case_when(
    gaze == "centre" ~ "centre",
    gaze_quadrant == location_deviant ~ "deviant",
    gaze_quadrant %in% c(location_standard, location_standard_smaller, location_standard_larger) ~ gaze  # standard
    ))
```

### without binning
```{r}
#within subject average
timecourse_within_subject <- data_time_course %>%
  filter(!is.na(gaze_is_on_deviant)) %>%
  group_by(
    participantid,
    event,
    shape,
    samplecount_trial             
  ) %>%
  summarise(
    pld_within_subj = mean(gaze_is_on_deviant, na.rm = TRUE),
    .groups = "drop"
  )
#grand average
timecourse_group <- timecourse_within_subject %>%
  group_by(
    event,
    shape,
    samplecount_trial
  ) %>%
  summarise(
    mean_pld = mean(pld_within_subj, na.rm = TRUE),
    se_pld = sd(pld_within_subj, na.rm = TRUE) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

```
### plot (without binning)
```{r}
p_withoutbinning<-ggplot(timecourse_group,
       aes(x = samplecount_trial, y = mean_pld, color = shape, fill = shape)) +
  geom_line() +
  geom_ribbon(aes(ymin = mean_pld - se_pld,
                  ymax = mean_pld + se_pld),
              alpha = 0.2, color = NA) +
  geom_hline(yintercept = 0.25, linetype = "dashed") +
  facet_grid(shape ~ event) +
  ylim(0, 0.75) +
  theme_classic() +
  labs(
    #title = "Timecourse of proportion looking at deviant across participants",
    subtitle = "X axis not binned",
    x = "Time (samples)",
    y = "Proportion looking at deviant"
  )+
  theme(
    legend.position = "none")
p_withoutbinning

```



### Bin the time axis 
```{r}
bin_width <- 5 #controls the smoothing of the graph

data_binned <- data_time_course %>%
  filter(
    validtrial == 1,
    ppt_included == 1,
    event %in% c("shapes_on_1", "shapes_on_2")
  ) %>%
  mutate(
    time_bin = ceiling(samplecount_trial / bin_width)
  )

```

### PLTD (Proportion looking time at deviant) per bin per trial
```{r}
pld_trial_bin <- data_binned %>%
  group_by(
    participantid, shape, event, block,
    trialnumber_withinblock, time_bin
  ) %>%
  summarise(
    pld_bin = mean(gaze_is_on_deviant, na.rm = TRUE),
    .groups = "drop"
  )

```
### Average within participant (per condition)
```{r}
pld_participant <- pld_trial_bin %>%
  group_by(participantid, shape, event, time_bin) %>%
  summarise(
    pld_participant = mean(pld_bin, na.rm = TRUE),
    .groups = "drop"
  )
#split by block
pld_participant_block <- pld_trial_bin %>%
  group_by(participantid, shape, block, event, time_bin) %>%
  summarise(
    pld_participant = mean(pld_bin, na.rm = TRUE),
    .groups = "drop"
  )


```
### average across participants (group level)
```{r}
pld_group <- pld_participant %>%
  group_by(shape, event, time_bin) %>%
  summarise(
    mean_pld = mean(pld_participant, na.rm = TRUE),
    se_pld   = sd(pld_participant, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

pld_group_block <- pld_participant_block %>%
  group_by(shape, block, event, time_bin) %>%
  summarise(
    mean_pld = mean(pld_participant, na.rm = TRUE),
    se_pld   = sd(pld_participant, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )


```
### plot
```{r}
p_withbinning <- ggplot(pld_group,
       aes(x = time_bin * bin_width, y = mean_pld, color = shape, fill = shape)) +
  geom_hline(yintercept = 0.25, linetype = "dashed") +
  geom_ribbon(
    aes(ymin = mean_pld - se_pld, ymax = mean_pld + se_pld),
    alpha = 0.2, color = NA
  ) +
  geom_line() +
  facet_grid(shape ~ event) +
  ylim(0, 0.75) +
  labs(
    subtitle = "X axis samples binned with bin length = 5",
    x = "Time (samples)",
    y = "Proportion looking at deviant")+
  theme_classic()+
  theme(
    legend.position = "none")

p_withbinning


```

```{r}
combined_timecourse <- ggarrange(
  p_withoutbinning,
  p_withbinning,
  ncol = 2, nrow = 1,
  labels = c("A", "B")
)

final_timecourse_figure <- annotate_figure(
  combined_timecourse,
  top = text_grob(
    "Timecourse of proportion looking at deviant across participants",
    face = "bold",
    size = 14
  )
)

final_timecourse_figure

ggsave(
  filename = "intruder_step006_timecourse.png",
  plot = final_timecourse_figure,
  width = 18,       # in inches
  height = 12,       # adjust as needed
  dpi = 300         # publication quality
)
```

### plot split across blocks
```{r}
#options(bitmapType = "cairo")

p_withbin_splitblock <- ggplot(pld_group_block,
       aes(x = time_bin * bin_width,
           y = mean_pld,
           color = shape,
           fill = shape)) +
  geom_hline(yintercept = 0.25,
             linetype = "dashed",
             color = "gray40") +
  geom_ribbon(
    aes(ymin = mean_pld - se_pld,
        ymax = mean_pld + se_pld),
    alpha = 0.2,
    color = NA
  ) +
  geom_line(size = 1) +
  facet_grid(event + shape ~ block) +
  scale_x_continuous(
    limits = c(0, 120),
    expand = c(0, 0)
  ) +
  scale_y_continuous(limits = c(0, 0.75)) +
  theme_classic(base_size = 15) +
  labs(
    title = "Time course of proportion looking to the deviant",
    subtitle = "Averaged across participants; split by block; Bin length = 5 samples",
    x = "Time (samples)",
    y = "Proportion looking to deviant",
    color = "Shape",
    fill = "Shape"
  ) +
  theme(
    legend.position = "none")

p_withbin_splitblock

# ggsave(
#   filename = "intruder_step006_timecoursebyblock.png",
#   plot = p_withbin_splitblock,
#   width = 14,       # in inches
#   height = 18,       # adjust as needed
#   dpi = 300         # publication quality
# )

```

