---
title: "step1_compute_gaze_data"
date: "2025-07-18"
---

This script does the following:

1. reads in & concatenantes individual data files into one matrix,

2. computes average gaze coordinates,

3. computes gaze location within specified AOIs,

4. saves the data w/ gaze location into one file `*.RData` that can be used for analysis.

**Input** files:

'[sub-]intruder_eyeData.csv'

from /data/raw/

**Output** files:

`data_output_step01.RData`

saved in /data/processed/

# Raw data file structure 

In SHAPES study, the raw data files contain the following columns. Some of the colum names are ammended below and saved in `aux_clean`.

`timestamp` eye tracker timestamp

`participantID` participant ID as assigned by the experimenter 

`ID` randomization list id, renamed below to **`listid`**

`expOrder` block id [range 1 to 4], renamed belwo to **`block`**


# Setup

## Libraries

```{r}
# Import the necessary libraries
library(readxl)
library(tidyverse)
library(rstatix)
```

## Experiment selection

```{r}
exp <- 'E1'

if (exp == "E1") {
  samplecountmin <- 115
  samplecountmax <- 125
  filename_randomization <- '20250708_random_intruder.csv'
}else {
  stop("Unknown experiment")
}
```

## Paths

```{r}
# Construct the paths
path_data_raw <- file.path(getwd(), paste0("data/raw_", exp))
path_data_processed <- file.path(getwd(), paste0("data/processed_", exp))
path_rand <- file.path(getwd(), "randomization")
path_wd <- file.path(getwd())
```

## Load data [eyetr]

```{r}
# Set the working directory
setwd(path_data_raw)

# Get file names of individual data files
tmp <- list.files(pattern = "intruder_eyeData.csv$")

# Read each csv file into a list
listData = lapply(tmp, read.csv)

# Concatenate the data in each file into one combined data frame
aux = do.call(rbind.data.frame, listData)
```


```{r}
rm(listData)
```

### Rename columns

```{r}
aux_clean <- aux %>% 
  rename(participantid = participantID, block = expOrder, listid = ID, trialnumber_withinblock = trialNumber) %>%
  group_by(participantid,block,trialnumber_withinblock,event) %>%
  mutate(samplecount_trial = row_number()) %>%
  ungroup()

str(aux_clean)
```

## Load data [randomization]

```{r}
setwd(path_rand)
randomization = read.csv(filename_randomization)
randomization <- randomization %>%
  rename(trialnumber_withinblock = trialNumber)
```

```{r}
str(randomization)
head(randomization)
```
```{r}
if (exp == 'E1')
  randomization <- randomization %>%
    mutate(across(
      .cols = c(location_deviant,location_standard,location_standard_smaller,location_standard_larger),
      .fns = ~ if_else(listid > 10 & shape %in% c('car','duck'), . +1, .)
    ))
```


## Merge data [eyetr, randomization]

```{r}
data_raw <- merge(aux_clean,randomization, by = c("listid","block","trialnumber_withinblock"))
```

```{r}
rm(aux_clean)
```

## Load coordinates

### Screen

Read in the screen scoordinates.

```{r}
setwd(path_wd)
source("coordinates_screen.txt")
```

### AOIs

Read in the coordinates of different AOIs.

```{r}
setwd(path_wd)
source("coordinates_aois.txt")
```

## Define functions

### average()

average() computes the average (x,y) coordinates of gaze

```{r}
average <- function(lx, rx, ly, ry){
    if (is.na(lx) & is.na(ly) & is.na(rx) & is.na(ry)) return(NA)
    else if (is.na(lx) & is.na(ly)) return (c(rx, ry))
    else if (is.na(rx) & is.na(ry)) return (c(lx, ly))
    else return (c((lx + rx)/2, (ly + ry)/2))
}
#if data from only one eye is available, we use that; if both, then calculate average
#average x coordinates and y
```

### gazeDirection()

```{r}
gazeDirection <- function(x, y){
    if (is.na(x) | is.na(y)) return("nogaze")
    else if (between(y, aoi_top_lower, aoi_top_upper)){
        if(between(x, aoi_left_west, aoi_left_east)) return("topleft_1")
        else if(between(x, aoi_right_west, aoi_right_east)) return("topright_2")
    }
    else if (between(y, aoi_bottom_lower, aoi_bottom_upper)){
        if(between(x, aoi_left_west, aoi_left_east)) return("bottomleft_3")
        else if(between(x, aoi_right_west, aoi_right_east)) return("bottomright_4")
    }
    else if (between(y, -0.075, 0.075) & between(x, -0.075, 0.075)) return("centre")
    else if (abs(x) > screen_xMax | abs(y) > screen_yMax) return("offscreen")
    return("onscreen_elsewhere")
}
#checks whether gaze point is within an AOI and return which one
```


# EVENT DURATION [timing check for quality control]

```{r}
data_raw %>%
  ungroup() %>%
  distinct(event)
```

```{r}
# Check the length (in samples) of events per trial
aux_event_duration <- data_raw %>%
    group_by(participantid, event, listid, block, trialnumber_withinblock) %>%
    summarise(samplecount = n()) %>%
  ungroup()
```

```{r}
aux_event_duration
```

## Plot

```{r}
aux_event_duration %>%
    ggplot(aes(x = trialnumber_withinblock, y = samplecount, color = as.factor(event))) +
    geom_jitter() 
```

## Descriptive statistics

TO DO: compute descriptive stats on event durations

- only central attention getter can have varied event duration but rest of the events were fixed time so shouldn't - find the violations 

```{r}
aux_event_duration %>%
  filter(event %in% c("shapes_on_1", "shapes_on_2")) %>%
  group_by(participantid, event) %>%
  get_summary_stats(samplecount)
```


```{r}
aux_violations_event_duration <- aux_event_duration %>%
  filter(
    (event %in% c("shapes_on_1", "shapes_on_2")) 
  ) %>%
  mutate(violation_event_duration = ifelse((samplecount < samplecountmin | samplecount > samplecountmax),TRUE,FALSE)) %>%
  select(participantid, block, trialnumber_withinblock, event, samplecount, violation_event_duration) %>%
  rename(totalsamples_per_event = samplecount)
```

```{r}
aux_violations_event_duration %>%
  filter(violation_event_duration == TRUE) %>%
  group_by(participantid,block,trialnumber_withinblock) %>%
  count()
```

```{r}
# Count total trials and valid trials
valid_event_summary <- aux_violations_event_duration %>%
  group_by(participantid, block, trialnumber_withinblock) %>%
  summarise(
    valid = all(!violation_event_duration),   # TRUE if no violation in the event
    .groups = "drop"
  ) %>%
  summarise(
    n_total_trials = n(),
    n_valid_trials = sum(valid),
    prop_valid = n_valid_trials / n_total_trials * 100
  )

valid_event_summary
```

```{r}
# Merge back into main data
data_raw_with_event_length <- merge(data_raw, aux_violations_event_duration, by = c("participantid", "event", "block", "trialnumber_withinblock")) %>%
  arrange(participantid,block,trialnumber_withinblock,timestamp)
```

# COMPUTE GAZE

## Average gaze coordinates

```{r}
data_output_step01 <- data_raw_with_event_length %>%
  rowwise %>%
  mutate(avgx = average(lx, rx, ly, ry)[1], 
         avgy = average(lx, rx, ly, ry)[2]) 
```

## Gaze location w/i AOIs

```{r}
data_output_step01 <- data_output_step01 %>% 
        rowwise %>% 
        mutate(gaze = factor(gazeDirection(avgx, avgy),
                             levels = c("topleft_1", "topright_2", 
                                        "bottomleft_3", "bottomright_4","centre",
                                        "onscreen_elsewhere",
                                        "offscreen","nogaze")))
```

### Check

```{r}
unique(data_output_step01$gaze)
```

```{r}
data_output_step01 %>% group_by(gaze) %>%
  filter(event %in% c("shapes_on_1", "shapes_on_2")) %>%
  count()
```

# Preview output data structure

```{r}
data_output_step01 <- dplyr::ungroup(data_output_step01)

str(data_output_step01)
```


# Save data

```{r}
setwd(path_data_processed)
save(
  data_output_step01,
  file = sprintf("data_output_step01_%s.RData", exp)
)
```

