---
title: "step2_compute_inclusion_criteria"
date: "2025-07-18"
---

Overview
This script:
1. Loads processed gaze data (step01)
2. Computes trial-level gaze quality metrics
3. Applies trial and participant inclusion criteria
4. Merges inclusion flags back into the dataset
5. Saves final dataset for analysis

# Setup

## Libraries

```{r}
# Import the necessary libraries
library(readxl)
library(tidyverse)
library(rstatix)
library(ggforce)
```

## Experiment selection

```{r}
exp <- 'E1' #or E2

if (exp == "E1") {
  samplecountmin <- 115
  samplecountmax <- 125
  filename_randomization <- '20250708_random_intruder.csv'
}else {
  stop("Unknown experiment")
}
```

## Paths

```{r}
# Construct the paths
path_data_processed <- file.path(getwd(), paste0("data/processed_", exp))
path_wd <- file.path(getwd())
```

## Load data [data_output_step01]

```{r}
setwd(path_data_processed)
load(sprintf("data_output_step01_%s.RData", exp))
```
## Load coordinates

### Screen

Read in the screen coordinates.

```{r}
setwd(path_wd)
source("coordinates_screen.txt")
```

### AOIs

Read in the coordinates of different AOIs.

```{r}
setwd(path_wd)
source("coordinates_aois.txt")
```

## Define AOIs for plotting

```{r}
aoi_left_bottom <- geom_rect(aes(xmin = aoi_left_west, xmax = aoi_left_east,
                             ymin = aoi_bottom_upper, ymax = aoi_bottom_lower),
                             fill = 'white', color = "darkcyan", alpha = 0, linewidth = 1) 

aoi_right_bottom <- geom_rect(aes(xmin = aoi_right_west, xmax = aoi_right_east,
                             ymin = aoi_bottom_upper, ymax = aoi_bottom_lower),
                             fill = 'white', color = "darkcyan", alpha = 0, linewidth = 1)

aoi_left_top <- geom_rect(aes(xmin = aoi_left_west, xmax = aoi_left_east,
                             ymin = aoi_top_lower, ymax = aoi_top_upper),
                             fill = 'white', color = "darkgreen", alpha = 0, linewidth = 1) 

aoi_right_top <- geom_rect(aes(xmin = aoi_right_west, xmax = aoi_right_east,
                             ymin = aoi_top_lower, ymax = aoi_top_upper),
                             fill = 'white', color = "darkgreen", alpha = 0, linewidth = 1) 

aoi_attention_getter <- geom_rect(aes(xmin = -.075, xmax = .075,
                             ymin = -.075, ymax = .075),
                             fill = 'white', color = "magenta", alpha = 0, linewidth = 1) 
```

# Explore

## Gaze [w/i time window]

### Preview 1 participant

```{r}
xCenter <- 0.6
options(repr.plot.width = 8, repr.plot.height = 20)

data_output_step01 %>% 
  filter(participantid == 101) %>% 
  filter(event %in% c('shapes_on_1','shapes_on_2')) %>%
  ggplot(aes(x = avgx, y = avgy, color = event)) +
   theme_classic() +
   coord_cartesian(xlim = c(-8/9, 8/9), 
                   ylim = c(-0.5, 0.5),
                   expand = FALSE) +
                     theme(axis.line = element_blank(),
                          axis.text.x = element_blank(),
                          axis.text.y = element_blank(),
                          axis.ticks = element_blank(),
                          axis.title.x = element_blank(),
                          axis.title.y = element_blank(),
                          panel.border = element_rect(colour = "black", fill = NA, size = 5)) +
                    aoi_left_bottom + aoi_left_top +
                    aoi_right_bottom + aoi_right_top + 
                    aoi_attention_getter +
                    geom_point()
```
### Preview all participants
```{r}

selected_events <- c("shapes_on_1", "shapes_on_2")

p <- data_output_step01 %>%
  filter(event %in% selected_events) %>%
  ggplot(aes(x = avgx, y = avgy, color = event)) +
  geom_point(alpha = 0.6, size = 0.8) +
  theme_classic() +
  coord_cartesian(xlim = c(-8/9, 8/9),
                  ylim = c(-0.5, 0.5),
                  expand = FALSE) +
  facet_wrap_paginate(~ participantid, ncol = 2, nrow = 2, page = 1) +  # shows 4 participants per page
  theme(
    axis.line = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, size = 1),
    legend.position = "bottom"
  ) +
  aoi_left_bottom + aoi_left_top +
  aoi_right_bottom + aoi_right_top +
  aoi_attention_getter +
  labs(title = "Gaze positions across participants (4 per page)",
       color = "Event")

n_pages <- n_pages(p)
for (i in seq_len(n_pages)) {
  print(p + facet_wrap_paginate(~ participantid, ncol = 2, nrow = 2, page = i))
}

```

## Gaze [timecourse] (preview 1 participant for 1 block)

```{r}
data_output_step01 %>% 
  filter(participantid == 101, block == 3) %>% 
  filter(event %in% c('shapes_on_1','shapes_on_2')) %>% 
  ggplot(aes(x = samplecount_trial, y = avgy),) + 
  theme_classic() + 
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 1)) + 
  #xlim(0,125) + 
  geom_point(size = .5) + 
  geom_point(aes(y = avgx), size = .5) + 
  geom_line(color = "blue") + #blue line for vertical avgy
  geom_line(aes(y = avgx), color = "red") + #red line for horizontal avgx
  facet_grid(event ~ trialnumber_withinblock)
```

## Gaze [trajectory w/ AOIs] (preview 1 participant for 1 block)

```{r}
data_output_step01 %>% 
  filter(participantid == 101, block == 1) %>% 
  filter(event %in% c('shapes_on_1','shapes_on_2')) %>%
  ggplot(aes(x = avgx, y = avgy, group = interaction(event, trialnumber_withinblock))) +
    geom_path(aes(color = samplecount_trial), size = 0.7) +  
    geom_point(size = 0.8, alpha = 0.6) +                    
    scale_color_viridis_c(option = "plasma", direction = 1) + # color = time progression
    theme_classic() +
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 1),
          aspect.ratio = 1,                                   
          strip.text = element_text(size = 8)) +
    coord_cartesian(xlim = c(-8/9, 8/9), ylim = c(-0.5, 0.5)) +       
    facet_grid(event ~ trialnumber_withinblock) +
    labs(x = "Horizontal gaze (avgx)", 
         y = "Vertical gaze (avgy)", 
         color = "Time (sample index)",
         title = "Gaze trajectory across trials and events") +
  aoi_left_bottom + aoi_left_top + 
  aoi_right_bottom + aoi_right_top + 
  aoi_attention_getter

```
# Attendance, inclusion

### Compute attendance [all events]

A sample is valid if we recorded gaze on the screen.

A sample is not valid if we did not recored gaze on the screen. This could be because we did not record gaze at all or we recorded gaze off-screen.

```{r}
aux_attendance_all_events <- data_output_step01 %>%
  group_by(participantid,block,trialnumber_withinblock,event,shape,location_deviant,violation_event_duration,totalsamples_per_event) %>%
  summarise(validsamples = sum(gaze %in% c("topleft_1", "topright_2",
                                           "bottomleft_3","bottomright_4","centre",
                                           "onscreen_elsewhere"), na.rm = TRUE),
            samples_shape_aois = sum(gaze %in% c("topleft_1", "topright_2",
                                           "bottomleft_3","bottomright_4"), na.rm = TRUE),
            invalidsamples = sum(gaze %in% c("offscreen","nogaze"), na.rm = TRUE),
  # fixation_ok check (≥18 consecutive AOI samples)
      fixation_ok = {
        aoi_vec <- gaze %in% c("topleft_1", "topright_2",
                               "bottomleft_3", "bottomright_4")
        runs <- rle(aoi_vec) #rle() = run-length encoding — Finds consecutive runs of same value
        any(runs$values == TRUE & runs$lengths >= 18)
      },
      .groups = "drop"
    ) %>%
  mutate(proponscreen = validsamples/(validsamples+invalidsamples)) 

str(aux_attendance_all_events)
head(aux_attendance_all_events,4)
```

### Compute trial inclusion [selected events]

*Intruder task*. A trial is valid iff it contains:
- selected event durations are valid
- a minimum 50% of valid samples,
- a minimum 1 fixation (i.e. 300 ms, i.e. >=18 samples) 

```{r}
selected_events = c('shapes_on_1', 'shapes_on_2')
```


```{r}
aux_duration_issues <- aux_attendance_all_events %>%
  filter(event %in% selected_events) %>%
  select(-validsamples,-invalidsamples, -violation_event_duration, samples_shape_aois) %>%
  pivot_wider(names_from = 'event', values_from = c(proponscreen,totalsamples_per_event,fixation_ok,samples_shape_aois)) %>%
  mutate(
    prop_onscreen_gaze_shapes_still_overall = (proponscreen_shapes_on_1+proponscreen_shapes_on_2)/2,
    fixation_ok_overall = fixation_ok_shapes_on_1 | fixation_ok_shapes_on_2,
    trialclassification = case_when(
      totalsamples_per_event_shapes_on_1 > samplecountmax |
      totalsamples_per_event_shapes_on_2 > samplecountmax ~ "trial_too_long",
      totalsamples_per_event_shapes_on_1 < samplecountmin |
      totalsamples_per_event_shapes_on_2 < samplecountmin ~ "trial_too_short",
      totalsamples_per_event_shapes_on_1 <= samplecountmax &
      totalsamples_per_event_shapes_on_2 <= samplecountmax &
      totalsamples_per_event_shapes_on_1 >= samplecountmin &
      totalsamples_per_event_shapes_on_2 >= samplecountmin ~ "trial_duration_okay",
    )
    )
```

```{r}
aux_duration_issues %>%
  group_by(trialclassification) %>%
  count()
```

```{r}
aux_duration_issues_count <- aux_duration_issues %>%
  group_by(participantid,trialclassification) %>%
  count() %>%
   pivot_wider(
    names_from = trialclassification, values_from = n,
    values_fill = 0   # fill missing combinations with 0
  ) %>%
  mutate(
    trial_duration_out_of_bounds =
      rowSums(across(any_of(c("trial_too_long", "trial_too_short"))))
  )
```

#### Exclute ppts with more than 2 trials with invalid duration

```{r}
aux_duration_issues_count %>%
  filter(trial_duration_out_of_bounds > 2) 
```

```{r}
aux_pptexclude_duration <- aux_duration_issues_count %>%
  filter(trial_duration_out_of_bounds > 2) %>%
  pull(participantid)
```

#### Exclute ppts with side biases

TO DO before preregistration: specify the criteria for side bias.For, now it is by individual observation of data and experimenter

```{r}
aux_pptexclude_sidebias <- if (exp == "E1") c(122) else NULL
```

#### Continued

```{r}
aux_attendance_shapes_still <- aux_attendance_all_events %>%
  filter(event %in% selected_events) %>%
  select(-validsamples,-invalidsamples, -violation_event_duration, samples_shape_aois) %>%
  pivot_wider(names_from = 'event', values_from = c(proponscreen,totalsamples_per_event,fixation_ok,samples_shape_aois)) %>%
  mutate(
    prop_onscreen_gaze_shapes_still_overall = (proponscreen_shapes_on_1+proponscreen_shapes_on_2)/2,
    fixation_ok_overall = fixation_ok_shapes_on_1 | fixation_ok_shapes_on_2,
    validtrial = ifelse(
      proponscreen_shapes_on_1 >= .5 & 
        proponscreen_shapes_on_2 >= .5 &
        (samples_shape_aois_shapes_on_1 > 18 | samples_shape_aois_shapes_on_2 > 18) &
        location_deviant > 0 &
        totalsamples_per_event_shapes_on_1 <= samplecountmax & totalsamples_per_event_shapes_on_2 <= samplecountmax &
        totalsamples_per_event_shapes_on_1 >= samplecountmin & totalsamples_per_event_shapes_on_2 >= samplecountmin &
        fixation_ok_overall == TRUE,
      1,
      0))

str(aux_attendance_shapes_still)
head(aux_attendance_shapes_still,50)
aux_ppt <- aux_attendance_shapes_still %>% filter(participantid == 126)
```

### Compute participant inclusion

```{r}
aux_valid_trial_participants <- aux_attendance_shapes_still %>%
  group_by(participantid,shape) %>%
  summarise(sumvalidtrials = sum(validtrial)) %>%
  pivot_wider(names_from = shape, values_from = sumvalidtrials, names_glue = "sumvalidtrials_{shape}") %>%
  mutate(across(everything(), ~ replace(., is.na(.), 0))) %>%
  mutate(sumvalidtrials_overall = rowSums(across(starts_with("sumvalidtrials_")), na.rm = TRUE)) %>%
  mutate(sumvalidtrials_object = sumvalidtrials_duck + sumvalidtrials_car) %>%
  mutate(ppt_included = if_else(
    if_any(
      c(starts_with("sumvalidtrials_"), -sumvalidtrials_car, -sumvalidtrials_duck),
      ~ . == 0
    )
    | participantid %in% aux_pptexclude_duration # [7/12/25] exclude ppts who had more 2 trials with invalid duration
    | participantid %in% aux_pptexclude_sidebias, # [7/12/25] exclude ppts with side biases
    0L, 1L
  )) 

aux_valid_trial_participants
```

## Merge data [eyetr, attendance]

```{r}
aux <- merge(data_output_step01,aux_attendance_shapes_still,
             by = c('participantid','block','trialnumber_withinblock','shape','location_deviant'))
```

## Merge data [eyetr, validtrials]

```{r}
data_output_step02 <- merge(aux,aux_valid_trial_participants,by = c('participantid'))
```

```{r}
str(data_output_step02)
head(data_output_step02)
```
## Print | count included ppts

```{r}
data_output_step02 %>%
  filter(ppt_included == 1) %>%
  distinct(participantid) %>%
  pull(participantid) %>%
  cat(sep = ", ")
```


```{r}
data_output_step02 %>%
  filter(ppt_included == 1) %>%
  distinct(participantid) %>%
  count()
```

## Print | count excluded ppts

```{r}
data_output_step02 %>%
  filter(ppt_included == 0) %>%
  distinct(participantid) %>%
  pull(participantid) %>%
  cat(sep = ", ")
```

```{r}
data_output_step02 %>%
  filter(ppt_included == 0) %>%
  distinct(participantid) %>%
  count()
```

## Print included lists

```{r}
data_output_step02 %>%
  filter(ppt_included == 1) %>%
  distinct(listid) %>%
  pull(listid) %>%
  cat(sep = ", ")
```
```{r}
data_output_step02 %>%
  filter(ppt_included == 1) %>%
  distinct(participantid, listid)

```


# Save data 

```{r}
setwd(path_data_processed)
save(
  data_output_step02,
  file = sprintf("data_output_step02_%s.RData", exp)
)
```

