---
title: "Step4_preliminiary_sanity_checks"
date: "2025-07-18"
---

Overview
This script:
1. Checks gaze distribution (across quadrants)
2. Examines gaze allocation by each stimuli on a single display (standard, small standard, large standard and deviant, i.e. also checks for size manipulations)
3. Tests for potential color biases
4. Produces sanity-check plots

# Setup

## Libraries

```{r}
# Import the necessary libraries
library(readxl)
library(tidyverse)
library(rstatix)
library(ggforce)
library(ggpubr)
library(scales)
library(afex)
library(emmeans)
```

## Paths

```{r}
exp <- 'E1' 
# Construct the paths
path_data_processed <- file.path(getwd(), paste0("data/processed_", exp))
path_wd <- file.path(getwd())
```

## Load data [data_output_step02]

```{r}
setwd(path_data_processed)
load(sprintf("data_output_step02_%s.RData", exp))
```
```{r}
data_output_step02 <- data_output_step02 %>%
  mutate(shape_w_duck_car = shape) %>%
  mutate(shape = ifelse(shape %in% c('duck','car'),'object',shape))

#for plotting purpose
if (exp == "E1") {
  desired_shape_order <- c(
    "object",
    "equilateral_triangle",
    "isosceles_triangle",
    "square",
    "rhombus",
    "pentagon"
  )
} 
```

##Proportion of gaze per quadrant

(computed first at participant-level then group level)

```{r}
aoi_quadrants <- c("topleft_1", "topright_2", "bottomleft_3", "bottomright_4")

# Participant-level proportion looking per quadrant
participant_gaze_summary <- data_output_step02 %>%
  filter(validtrial == 1, ppt_included == 1, 
         event %in% c("shapes_on_1", "shapes_on_2")) %>%
  filter(!gaze %in% c("offscreen", "nogaze")) %>%
  group_by(participantid, shape, block, gaze) %>% 
  summarise(gaze_count = n(), .groups = "drop") %>%
  group_by(participantid, shape, block) %>%
  mutate(total_onscreen = sum(gaze_count)) %>%
  ungroup() %>%
  filter(gaze %in% aoi_quadrants) %>%
  pivot_wider(names_from = gaze, values_from = gaze_count, values_fill = 0) %>%
  mutate(
    total_shapes_aoi = rowSums(across(all_of(aoi_quadrants))),
    prop_topleft_1 = topleft_1 / total_shapes_aoi,
    prop_topright_2 = topright_2 / total_shapes_aoi,
    prop_bottomleft_3 = bottomleft_3 / total_shapes_aoi,
    prop_bottomright_4 = bottomright_4 / total_shapes_aoi
  ) %>%
  pivot_longer(
    cols = starts_with("prop_"),
    names_to  = "aoi",
    values_to = "prop",
    names_prefix = "prop_"
  )

participant_gaze_summary %>%
  group_by(aoi) %>%
  get_summary_stats(prop, type = "full")

#stats
aoi_means <- participant_gaze_summary %>%
  filter(!is.na(prop)) %>%
  group_by(participantid, aoi) %>%
  summarise(prop_mean = mean(prop, na.rm = TRUE), .groups = "drop")

anova_results <- aoi_means %>%
  anova_test(
    dv = prop_mean,
    wid = participantid,
    within = aoi
  )
anova_results

pairwise_results <- aoi_means %>%
  pairwise_t_test(
    prop_mean ~ aoi,
    paired = TRUE,
    p.adjust.method = "BH"
  )

pairwise_results


#Plot
p_quadrant <- participant_gaze_summary %>%
  group_by(participantid, aoi) %>%
  summarise(m_prop = mean(prop, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = aoi, y = m_prop, fill = aoi)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 1.5) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "gray40") +
  stat_summary(fun = mean, geom = "point", size = 4, color = "tomato") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_classic(base_size = 8) +
  labs(
    title = "Participant-level Proportion of Gaze by Quadrant",
    x = "Screen Quadrant",
    y = "Mean Proportion of Looking"
  )+
  theme(
    legend.position = "none")

dodge_width <- 0.8

# Compute mean per participant per quadrant per shape
participant_gaze_shape_summary <- participant_gaze_summary %>%
  group_by(participantid, shape, aoi) %>%
  summarise(m_prop = mean(prop, na.rm = TRUE), .groups = "drop")

# Plot
p_quadrant_by_shape <- ggplot(participant_gaze_shape_summary, aes(x = aoi, y = m_prop, fill = aoi)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA, color = "black", position = position_dodge(width = dodge_width)) +
  geom_jitter(aes(color = aoi), alpha = 0.7, size = 1.5, position = position_dodge(width = dodge_width)) +
  stat_summary(fun = mean, geom = "point", size = 3, color = "tomato", position = position_dodge(width = dodge_width)) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "gray40") +
  facet_wrap(~shape) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_classic(base_size = 8) +
  labs(
    title = "Participant-level Proportion of Gaze by Quadrant and Shape",
    x = "Screen Quadrant",
    y = "Mean Proportion of Looking",
    fill = "Quadrant",
    color = "Quadrant"
  ) +
  theme(legend.position = "none",axis.text.x = element_text(angle = 45, hjust = 1))



```


##Proportion of gaze by shape size

```{r}
participant_gaze_shape_size <- data_output_step02 %>%
  filter(ppt_included == 1, validtrial == 1) %>%
  filter(event %in% c("shapes_on_1", "shapes_on_2"),
         !gaze %in% c("offscreen", "nogaze")) %>%
  mutate(
    # numeric quadrant from gaze name
    gaze_quadrant = case_when(
      gaze == "topleft_1" ~ 1,
      gaze == "topright_2" ~ 2,
      gaze == "bottomleft_3" ~ 3,
      gaze == "bottomright_4" ~ 4,
      TRUE ~ NA_real_
    ),
    # whether gaze matches deviant/standard positions
    match_deviant = gaze_quadrant == location_deviant,
    match_standard = gaze_quadrant == location_standard,
    match_standard_smaller = gaze_quadrant == location_standard_smaller,
    match_standard_larger = gaze_quadrant == location_standard_larger
  ) %>%
  group_by(participantid, shape) %>% 
  summarise(
    prop_look_deviant = mean(match_deviant, na.rm = TRUE),
    prop_look_standard = mean(match_standard, na.rm = TRUE),
    prop_look_standard_smaller = mean(match_standard_smaller, na.rm = TRUE),
    prop_look_standard_larger = mean(match_standard_larger, na.rm = TRUE),
    .groups = "drop"
  )
#reshape for plotting
gaze_shape_long <- participant_gaze_shape_size %>%
  pivot_longer(
    cols = starts_with("prop_look"),
    names_to = "stimuli_type",
    values_to = "proportion",
    names_prefix = "prop_look_"
  )
#group-level summary
gaze_shape_long %>%
  group_by(shape, stimuli_type) %>%
  summarise(
    mean_prop = mean(proportion, na.rm = TRUE),
    sd_prop = sd(proportion, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

#stats


# two-way repeated-measures ANOVA
anova_results <- aov_ez(
  id = "participantid",
  dv = "proportion",
  within = c("shape", "stimuli_type"),
  data = gaze_shape_long,
  factorize = TRUE,
  type = 3
)

anova_results

emm <- emmeans(anova_results, ~ stimuli_type | shape)
pairs(emm, adjust = "holm")  # corrected for multiple comparisons

#plot
dodge_width <- 0.8
p_shape_size <- gaze_shape_long %>% mutate(shape = factor(shape, levels = desired_shape_order)) %>%
  ggplot(aes(x = shape, y = proportion, fill = stimuli_type)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA, color = "black",position = position_dodge(width = dodge_width)) +
  geom_jitter(aes(color = stimuli_type), alpha = 0.7, size = 1.5, position = position_dodge(width = dodge_width)) +
  stat_summary(fun = mean, geom = "point", size = 3, color = "tomato", position = position_dodge(width = dodge_width)  # align with boxplot
  ) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "gray40") +
  facet_wrap(~stimuli_type) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_classic(base_size = 8) +
  labs(
    x = "Shape",
    y = "Proportion of Looking",
    fill = "Stimuli Type",
    color = "Stimuli Type",
    title = "Proportion of Gaze by Shape and Stimuli Type (Size)"
  ) +
  theme(legend.position = "none",axis.text.x = element_text(angle = 45, hjust = 1))

#collapsed plot across shapes


# Plot
p_location_collapsed <- gaze_shape_long %>% mutate(shape = factor(shape, levels = desired_shape_order)) %>%
ggplot(aes(x = shape, y = proportion, fill = shape)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA, color = "black", position = position_dodge(width = dodge_width)) +
  geom_jitter(aes(color = shape), alpha = 0.7, size = 1.5, position = position_dodge(width = dodge_width)) +
  stat_summary(fun = mean, geom = "point", size = 3, color = "tomato", position = position_dodge(width = dodge_width)) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "gray40") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_classic(base_size = 8) +
  labs(
    x = "Stimuli Type",
    y = "Proportion of Looking",
    fill = "Stimuli Type",
    color = "Stimuli Type",
    title = "Proportion of Gaze by shape collapsed across stimuli type"
  ) +
  theme(legend.position = "none",axis.text.x = element_text(angle = 45, hjust = 1))


```

##Proportion looking at deviant by color
```{r}
gaze_by_color <- data_output_step02 %>%
  filter(validtrial == 1, ppt_included == 1) %>%
  filter(event %in% c("shapes_on_1", "shapes_on_2")) %>%
  filter(shape != "object") %>%  # exclude 'object'
  # Keep only on-screen gazes
  filter(!gaze %in% c("offscreen", "nogaze")) %>%
  mutate(
    gaze_quadrant = case_when(
      gaze == "topleft_1" ~ 1,
      gaze == "topright_2" ~ 2,
      gaze == "bottomleft_3" ~ 3,
      gaze == "bottomright_4" ~ 4,
      TRUE ~ NA_real_
    ),
    match_deviant = gaze_quadrant == location_deviant
  ) %>%
  group_by(participantid, shape, color) %>% 
  summarise(
    prop_look_deviant = mean(match_deviant, na.rm = TRUE),
    .groups = "drop"
  )

# Group-level summary 
gaze_by_color_summary <- gaze_by_color %>%
  group_by(color) %>%
  summarise(
    mean_prop_look_deviant = mean(prop_look_deviant, na.rm = TRUE),
    sd_prop_look_deviant = sd(prop_look_deviant, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
gaze_by_color_summary

#stats
gaze_by_color_participant <- gaze_by_color %>%
  group_by(participantid, color) %>%
  summarise(prop_look_deviant = mean(prop_look_deviant), .groups = "drop")

anova_results <- gaze_by_color_participant %>%
  anova_test(
    dv = prop_look_deviant,
    wid = participantid,
    within = color
  )
anova_results


# Plot
p_color <- ggplot(gaze_by_color, aes(x = color, y = prop_look_deviant, fill = color)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA, color = "black") +
  geom_jitter(width = 0.15, alpha = 0.7, size = 2, aes(color = color)) +
  stat_summary(fun = mean, geom = "point", shape = 21, size = 4,
               fill = "tomato", color = "black", stroke = 0.8) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "gray40") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_classic(base_size = 8) +
  labs(
    title = "Proportion of Looking at Deviant by Shape Color",
    x = "Color",
    y = "Mean Proportion Looking",
    fill = "Color",
    color = "Color"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

## All plots
```{r}
final_figure <- ggarrange(
  p_quadrant, p_quadrant_by_shape,p_location_collapsed, p_shape_size, p_color,
  ncol =5 , nrow = 1, labels = c("A", "B", "C", "D", "E")
)
final_figure
```
```{r}
ggsave(
  filename = "intruder_step004_checks.png",
  plot = final_figure,
  width = 25,       # in inches
  height = 10,       # adjust as needed
  dpi = 300        
)
```