---
title: "step2_compute_screen_attendance"
date: "2025-07-18"
---

 Overview
 This script:
 1. Summarises valid trials per participant
 2. Computes on-screen gaze (attendance)
 3. Examines attendance across blocks and shapes
 4. Produces summary plots for reporting

# Setup

## Libraries

```{r}
# Import the necessary libraries
library(readxl)
library(tidyverse)
library(rstatix)
library(ggforce)
library(ggpubr)
library(scales)
```
```{r}
exp <- 'E1' 
```


## Paths

```{r}
# Construct the paths
path_data_processed <- file.path(getwd(), paste0("data/processed_", exp))
path_wd <- file.path(getwd())
```

## Load data [data_output_step02]

```{r}
setwd(path_data_processed)
load(sprintf("data_output_step02_%s.RData", exp))
```
```{r}
data_output_step02 <- data_output_step02 %>%
  mutate(shape_w_duck_car = shape) %>%
  mutate(shape = ifelse(shape %in% c('duck','car'),'object',shape))
```
```{r}
#for plotting purpose
if (exp == "E1") {
  desired_shape_order <- c(
    "object",
    "equilateral_triangle",
    "isosceles_triangle",
    "square",
    "rhombus",
    "pentagon"
  )
} 
```

#Attendance

##Number of valid trials

###overall stats
```{r}
data_output_step02 %>%
  filter(ppt_included == 1) %>% 
  distinct(participantid, sumvalidtrials_overall) %>%
  get_summary_stats(sumvalidtrials_overall, type = "full")
#plot
pA <- data_output_step02 %>%
  filter(ppt_included == 1) %>%
  distinct(participantid, sumvalidtrials_overall) %>%
  ggplot(aes(x = sumvalidtrials_overall)) +
  geom_histogram(binwidth = 1, fill = "#4C72B0", color = "black", alpha = 0.8) +
  theme_classic(base_size = 13) +
  labs(
    title = "A. Valid trials per infant: frequency distribution",
    x = "Number of valid trials",
    y = "Number of infants"
  )

```

###block-wise
```{r}
data_output_step02 %>%
  filter(ppt_included == 1, validtrial == 1) %>%
  distinct(participantid, block, shape, trialnumber_withinblock, .keep_all = TRUE) %>%
  group_by(participantid,block) %>%            # first, count per participant
  summarise(n_valid_trials = n(), .groups = "drop") %>%
  group_by(block) %>%                           # then, aggregate across participants
  get_summary_stats(n_valid_trials, type = "mean")
#plot
block_summary <- data_output_step02 %>%
  filter(ppt_included == 1, validtrial == 1) %>%
  distinct(participantid, block, shape, trialnumber_withinblock) %>%
  group_by(participantid, block) %>%
  summarise(n_valid_trials = n(), .groups = "drop")

pB <- block_summary %>%
  ggplot(aes(x = factor(block), y = n_valid_trials)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA, width = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 1.5) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, size = 3, fill = "tomato") +
  theme_classic(base_size = 13) +
  labs(
    title = "B. Number of Valid trials by block",
    x = "Block",
    y = "Number of valid trials"
  )

```
### valid trials per condition (shape)
```{r}
data_output_step02 %>%
  filter(ppt_included == 1, validtrial == 1) %>%
  distinct(participantid, block, shape, trialnumber_withinblock, .keep_all = TRUE) %>%
  group_by(participantid,shape) %>%            # first, count per participant
  summarise(n_valid_trials = n(), .groups = "drop") %>%
  group_by(shape) %>%                           # then, aggregate across participants
  get_summary_stats(n_valid_trials, type = "mean_sd")

#plot
shape_summary <- data_output_step02 %>%
  filter(ppt_included == 1, validtrial == 1) %>%
  distinct(participantid, block, shape, trialnumber_withinblock) %>%
  group_by(participantid, shape) %>%
  summarise(n_valid_trials = n(), .groups = "drop")

pC <- shape_summary %>%
  mutate(shape = factor(shape, levels = desired_shape_order)) %>%
  ggplot(aes(x = shape, y = n_valid_trials, fill = shape)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA, width = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 1.5) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, size = 3, fill = "tomato") +
  theme_classic(base_size = 13) +
  labs(
    title = "C. Number of valid trials by shape",
    x = "Shape",
    y = "Number of valid trials"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


```
##On-screen gaze during valid trials
```{r}
attendance_summary <- data_output_step02 %>%
  filter(ppt_included == 1, validtrial == 1) %>%
  group_by(participantid) %>%
  summarise(mean_proponscreen = mean((proponscreen_shapes_on_1 + proponscreen_shapes_on_2)/2,
                                     na.rm = TRUE)) %>%
  get_summary_stats(mean_proponscreen, type = "full")

attendance_summary

#plot
pD <- data_output_step02 %>%
  filter(ppt_included == 1, validtrial == 1) %>%
  group_by(participantid) %>%
  summarise(mean_proponscreen = mean((proponscreen_shapes_on_1 +
                                       proponscreen_shapes_on_2) / 2,
                                     na.rm = TRUE)) %>%
  ggplot(aes(y = mean_proponscreen, x = "")) +
  geom_boxplot(fill = "#55A868", alpha = 0.5, width = 0.6) +
  geom_jitter(width = 0.05, alpha = 0.6) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  theme_classic(base_size = 13) +
  labs(
    title = "D. Overall on-screen gaze",
    x = NULL,
    y = "Proportion on-screen"
  )

```

###screen attendance by blocks
```{r}
attendance_by_block <- data_output_step02 %>%
  filter(ppt_included == 1, validtrial == 1) %>%
  # Compute mean screen attendance per participant per block
  group_by(participantid, block) %>%
  summarise(mean_proponscreen = mean((proponscreen_shapes_on_1 + proponscreen_shapes_on_2)/2,
                                     na.rm = TRUE),
            .groups = "drop") %>%
  # Now get group-level summary (M, SD across participants)
  group_by(block) %>%
  get_summary_stats(mean_proponscreen, type = "mean_sd")

attendance_by_block

#plot
pE <- data_output_step02 %>%
  filter(ppt_included == 1, validtrial == 1) %>%
  group_by(participantid, block) %>%
  summarise(mean_proponscreen = mean((proponscreen_shapes_on_1 +
                                       proponscreen_shapes_on_2) / 2,
                                     na.rm = TRUE),
            .groups = "drop") %>%
  ggplot(aes(x = factor(block), y = mean_proponscreen)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA, width = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.6) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, size = 3, fill = "tomato") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  theme_classic(base_size = 13) +
  labs(
    title = "E. On-screen gaze by block",
    x = "Block",
    y = "Proportion on-screen"
  )

```
### screen attendance by shape
```{r}
attendance_by_shape <- data_output_step02 %>%
  filter(ppt_included == 1, validtrial == 1) %>%
  # Compute mean screen attendance per participant per shape
  group_by(participantid, shape) %>%
  summarise(mean_proponscreen = mean((proponscreen_shapes_on_1 + proponscreen_shapes_on_2)/2,
                                     na.rm = TRUE),
            .groups = "drop") 

attendance_by_shape %>%
# Now get group-level summary (M, SD across participants)
  group_by(shape) %>%
  get_summary_stats(mean_proponscreen, type = "mean_sd")

#stats
anova_results <- attendance_by_shape %>%
  anova_test(dv = mean_proponscreen,
             wid = participantid,
             within = shape)

anova_results

pairwise_result <- attendance_by_shape %>%
  pairwise_t_test(mean_proponscreen ~ shape,
                  paired = TRUE,
                  p.adjust.method = "holm")
pairwise_result

#plot
pF <- data_output_step02 %>%
  filter(ppt_included == 1, validtrial == 1) %>%
  group_by(participantid, shape) %>%
  summarise(
    mean_proponscreen = mean((proponscreen_shapes_on_1 +
                               proponscreen_shapes_on_2) / 2,
                             na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(shape = factor(shape, levels = desired_shape_order)) %>%
  ggplot(aes(x = shape, y = mean_proponscreen, fill = shape)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA, width = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 1.5) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, size = 3, fill = "tomato") +
  theme_classic(base_size = 13) +
  labs(
    title = "F. Screen Attendance per Shape (Participant-level)",
    subtitle = "Each point = mean screen attendance per participant",
    x = "Shape",
    y = "Proportion on-screen"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


```
##All plots
```{r}
final_row1 <- ggarrange(
  pA, pB, pC,
  ncol = 3,
  labels = c("A", "B", "C")
)

final_row2 <- ggarrange(
  pD, pE, pF,
  ncol = 3,
  labels = c("D", "E", "F")
)

final_figure <- ggarrange(
  ggarrange(
    text_grob("Valid trials", face = "bold", size = 14),
    final_row1,
    ncol = 1,
    heights = c(0.1, 1)
  ),
  ggarrange(
    text_grob("Screen attendance during valid trials", face = "bold", size = 14),
    final_row2,
    ncol = 1,
    heights = c(0.1, 1)
  ),
  nrow = 2
)

final_figure

```
```{r}
ggsave(
  filename = "intruder_step003_attendance.png",
  plot = final_figure,
  width = 18,       # in inches
  height = 12,       # adjust as needed
  dpi = 300        
)
```

