---
title: "step_05_main_measure_PLTD"
date: "2025-07-21"
---

Overview
This script computes and analyses the main measure - Proportion Looking Time at Deviant
1. Computes PLTD at trial level (dwell time on deviant AOI/ total dwell on all AOIs)
2. Aggregate to participant level
3. Create group level dataset - overall, by block (learning/fatigue), by event(early vs late display)
4. Statistical analysis

Note - chance level is at 0.25 (4 quadrants)
All analyses use valid trials and included participants only

# Setup

## Libraries

```{r}
# Import the necessary libraries
library(readxl)
library(tidyverse)
library(rstatix)
library(ggforce)
library(ggpubr)
library(patchwork)
```

## Experiment selection

```{r}
exp <- 'E1' 

#for plotting purpose
if (exp == "E1") {
  levels_labels <- c(
    "object",
    "equilateral_triangle",
    "isosceles_triangle",
    "square",
    "rhombus",
    "pentagon"
  )
}


```

## Paths

```{r}
# Construct the paths
path_data_processed <- file.path(getwd(), paste0("data/processed_", exp))
path_wd <- file.path(getwd())
```

## Load data [eyetr w/ gaze locations]

```{r}
setwd(path_data_processed)
load(sprintf("data_output_step02_%s.RData", exp))
```

```{r}
str(data_output_step02)
```

```{r}
data_output_step02 <- data_output_step02 %>%
  mutate(shape_w_duck_car = shape) %>%
  mutate(shape = ifelse(shape %in% c('duck','car'),'object',shape))
```

# MEASURE: prop_look_dev

## Individual [averages] 

Measure: proportion of looking at the deviant

### Collapsed across events

```{r}
aux_prop_look_dev <- data_output_step02 %>%
  filter(event %in% c('shapes_on_1','shapes_on_2')) %>%
  group_by(participantid,ppt_included,block,trialnumber_withinblock,shape,gaze,
           location_deviant,deviant_type,
           orientation,
           validtrial,sumvalidtrials_overall,
           proponscreen_shapes_on_1,proponscreen_shapes_on_2,shape_w_duck_car) %>% 
  count() %>%
  pivot_wider(names_from = gaze, values_from = n, values_fill = 0, names_glue = "gaze_{gaze}")%>%
  mutate(total_dwell_shapes = gaze_topleft_1 + gaze_topright_2 + gaze_bottomleft_3 + gaze_bottomright_4) %>%
  mutate(total_dwell_deviant = case_when(
    location_deviant == 1 ~ gaze_topleft_1,
    location_deviant == 2 ~ gaze_topright_2,
    location_deviant == 3 ~ gaze_bottomleft_3,
    location_deviant == 4 ~ gaze_bottomright_4)) %>% 
  mutate(prop_look_dev = (total_dwell_deviant/total_dwell_shapes))
```

### Split by event (shapes_on_1 vs shapes_on_2)

```{r}
aux_prop_look_dev_by_event <- data_output_step02 %>%
  filter(event %in% c('shapes_on_1','shapes_on_2')) %>%
  group_by(participantid,ppt_included,block,trialnumber_withinblock,shape,gaze,location_deviant,deviant_type,orientation,
           validtrial,sumvalidtrials_overall,event,
           proponscreen_shapes_on_1,proponscreen_shapes_on_2,shape_w_duck_car) %>%
  count() %>%
  pivot_wider(names_from = gaze, values_from = n, values_fill = 0, names_glue = "gaze_{gaze}")%>%
  mutate(total_dwell_shapes = gaze_topleft_1 + gaze_topright_2 + gaze_bottomleft_3 + gaze_bottomright_4) %>%
  mutate(total_dwell_deviant = case_when(
    location_deviant == 1 ~ gaze_topleft_1,
    location_deviant == 2 ~ gaze_topright_2,
    location_deviant == 3 ~ gaze_bottomleft_3,
    location_deviant == 4 ~ gaze_bottomright_4)) %>% 
  mutate(prop_look_dev = (total_dwell_deviant/total_dwell_shapes))
```

## Group [averages] 

### Overall
```{r}
aux_prop_look_dev_group <- aux_prop_look_dev %>%
  filter(validtrial == 1) %>%
  filter(ppt_included == 1) %>%
  group_by(participantid,shape) %>%
  summarise(m_prop_look_dev = mean(prop_look_dev, na.rm = T))
```

### Split by block

```{r}
aux_prop_look_dev_group_by_block <- aux_prop_look_dev %>%
  filter(validtrial == 1) %>%
  filter(ppt_included == 1) %>%
  group_by(participantid,shape,block) %>%
  summarise(m_prop_look_dev = mean(prop_look_dev, na.rm = T))
```

### Split by event [on_1 vs on_2]

```{r}
aux_prop_look_dev_group_by_event <- aux_prop_look_dev_by_event %>%
  filter(validtrial == 1) %>%
  filter(ppt_included == 1) %>%
  group_by(participantid,shape,event) %>%
  summarise(m_prop_look_dev = mean(prop_look_dev, na.rm = T))
```

```{r}
aux_prop_look_dev_group %>%
  filter(shape == 'square') %>%
  filter(m_prop_look_dev < .25)
```

```{r}
aux_prop_look_dev_group %>%
  filter(shape == 'square') %>%
  filter(m_prop_look_dev >= .25)
```

# PLOTS

## Group

### Boxplot [overall]

```{r}
p_overall <- aux_prop_look_dev_group %>% 
  mutate(shape = factor(shape, levels = levels_labels)) %>%
  ggplot(aes(y = m_prop_look_dev, x = shape, color = shape)) +
  geom_boxplot() +
  geom_point(position=position_dodge(width=.1)) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "gray40") +
  stat_summary(fun = mean, geom = "point", size = 4, color = "tomato") +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  ylim(0,0.6)

p_overall

ggsave(
  filename = "intruder_step005_mainpropdev.png",
  plot = p_overall,
  width = 18,       # in inches
  height = 12,       # adjust as needed
  dpi = 300         # publication quality
)
``` 


### Boxplot [split by event]

```{r}
p_event <- aux_prop_look_dev_group_by_event %>% 
  mutate(shape = factor(shape, levels = levels_labels)) %>%
  ggplot(aes(y = m_prop_look_dev, x = shape, color = shape)) +
  geom_boxplot() +
  geom_point(position=position_dodge(width=.1)) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "gray40") +
  stat_summary(fun = mean, geom = "point", size = 4, color = "tomato") +
  theme(legend.position = "none",axis.text.x = element_text(angle = 45, hjust=1)) +
  facet_wrap(. ~ event)
```

### Boxplot [blocks 1 vs 2 vs 3 vs 4]

```{r}
p_individual_blocks <- aux_prop_look_dev_group_by_block %>% 
  mutate(shape = factor(shape, levels = levels_labels)) %>%
  ggplot(aes(y = m_prop_look_dev, x = shape, color = shape)) +
  geom_boxplot() +
  geom_point(position=position_dodge(width=.1)) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "gray40") +
  stat_summary(fun = mean, geom = "point", size = 4, color = "tomato") +
  theme(legend.position = "none",axis.text.x = element_text(angle = 45, hjust=1)) +
  facet_wrap(. ~ block)

p_individual_blocks
```

### Boxplot block1-2 vs block 3-4

```{r}
p_block <- aux_prop_look_dev_group_by_block %>% 
  mutate(shape = factor(shape, levels = levels_labels)) %>%
  mutate(block_group = case_when(
      block %in% c(1, 2) ~ "Blocks 1-2",
      block %in% c(3, 4) ~ "Blocks 3-4"
    )) %>%
  ggplot(aes(y = m_prop_look_dev, x = shape, color = shape)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position=position_dodge(width=.1), size=0.75) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "gray40") +
  stat_summary(fun = mean, geom = "point", size = 1.8, color = "tomato") +
  theme(legend.position = "none",axis.text.x = element_text(angle = 45, hjust=1)) +
  facet_wrap(. ~ block_group)
```


```{r}
ggarrange(
  p_event, p_block,
  ncol = 1, nrow = 2, labels = c("A", "B")
)
```

## Individual

#### Boxplots
```{r}
aux_prop_look_dev %>% 
  filter(ppt_included == 1) %>%
  filter(validtrial == 1) %>%
  mutate(shape = factor(shape, levels = levels_labels)) %>%
  ggplot(aes(y = prop_look_dev, x = shape, color = shape)) +
  geom_boxplot() +
  geom_point(position=position_dodge(width=.1)) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "gray40") +
  stat_summary(fun = mean, geom = "point", size = 3, color = "tomato") +
  facet_wrap_paginate(~ participantid, ncol = 2, nrow = 2, page = 1) +  # shows 4 participants per page
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  facet_wrap(. ~ participantid)
```
#### Line
```{r}
aux_prop_look_dev %>% 
  filter(ppt_included == 1) %>%
  filter(validtrial == 1) %>%
  filter(shape != "object") %>%
  mutate(shape = factor(shape, levels = levels_labels)) %>%
  ggplot(aes(y = prop_look_dev, x = shape, color = shape)) +
  #geom_boxplot() +
  #geom_point(position=position_dodge(width=.1)) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "gray40") +
  stat_summary(fun = mean, geom = "point", size = 3, aes(color = shape)) +
  stat_summary(fun = mean, geom = "line", aes(group = 1), size = 1, color = "gray") +
  facet_wrap_paginate(~ participantid, ncol = 2, nrow = 2, page = 1) +  # shows 4 participants per page
  theme(legend.position = "none",axis.text.x = element_text(angle = 45, hjust=1)) +
  facet_wrap(. ~ participantid)
```


```{r}
aux_prop_look_dev %>% 
  filter(ppt_included == 1, validtrial == 1) %>%
  filter(shape != "object") %>%
  mutate(shape = factor(shape, levels = levels_labels)) %>%
  ggplot(aes(x = shape, y = prop_look_dev, group = participantid)) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "gray40") +
  stat_summary(fun = mean, geom = "line", aes(group = participantid), color = "gray70", size = 0.5) +
  stat_summary(fun = mean, geom = "line", aes(group = 1), color = "tomato", size = 1.5) + # group mean
  # add mean points per shape (group mean)
  stat_summary(fun = mean, geom = "point", size = 3, aes(color = shape)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y = "Proportion looking at the deviant", x = "Shape")
```
# STATISTICS

# Statistics: prop_look_dev
```{r}
aux_prop_look_dev_group %>%
  group_by(shape) %>%
  get_summary_stats(m_prop_look_dev, type = "mean_sd")
```

## Shapiro-Wilk test (normality assumption)

```{r}
aux_prop_look_dev_group %>%
  group_by(shape) %>%
  shapiro_test(m_prop_look_dev)
```

## Comparison against chance

```{r}
aux_prop_look_dev_group %>%
  group_by(shape) %>%
  t_test(m_prop_look_dev ~ 1, mu = .25)
```
## Pairwise comparisons of shapes

```{r}
aux <- aux_prop_look_dev_group %>% 
  filter(!shape %in% c('car', 'duck')) %>%
  convert_as_factor(participantid, shape) %>%
  ungroup()
```

```{r}
auxcheck <- aux %>%
  pivot_wider(values_from = m_prop_look_dev, names_from = 'shape') 
```

```{r}
aux %>%
  pairwise_t_test(m_prop_look_dev ~ shape, 
                  subject = "participantid",
                  paired = TRUE, p.adjust.method = "bonferroni")
```

## Anova

```{r}
res.aov <- anova_test(
  data = aux, dv = m_prop_look_dev, wid = participantid,
  within = c(shape)
  )
```

```{r}
res.aov
```

## Condition weighted means 
(each participant has multiple rows for each condition)
```{r}
aux_prop_look_dev_condition <- aux_prop_look_dev %>% 
  filter(validtrial == 1, ppt_included == 1) %>%
  group_by(participantid, shape, deviant_type, orientation, location_deviant) %>%
  summarise(m_prop_look_dev = mean(prop_look_dev, na.rm = TRUE),
            .groups = "drop")

```

```{r}
# Mean ± SD by shape
aux_prop_look_dev_condition %>%
  group_by(shape) %>%
  get_summary_stats(m_prop_look_dev, type = "mean_sd")

# Mean ± SD by deviant_type
aux_prop_look_dev_condition %>%
  group_by(shape,deviant_type) %>%
  get_summary_stats(m_prop_look_dev, type = "mean_sd")

# Mean ± SD by orientation
aux_prop_look_dev_condition %>%
  group_by(shape,orientation) %>%
  get_summary_stats(m_prop_look_dev, type = "mean_sd")

# Mean ± SD by deviant location
aux_prop_look_dev_condition %>%
  group_by(shape,location_deviant) %>%
  get_summary_stats(m_prop_look_dev, type = "mean_sd")
```
```{r}
# By deviant_type
anova_deviant_type <- aux_prop_look_dev_condition %>%
  group_by(shape) %>%
  anova_test(m_prop_look_dev ~ deviant_type)

anova_deviant_type %>% get_anova_table()

# By orientation
anova_orientation <- aux_prop_look_dev_condition %>%
  group_by(shape) %>%
  anova_test(m_prop_look_dev ~ orientation)

anova_orientation %>% get_anova_table()

# By location_deviant
anova_location <- aux_prop_look_dev_condition %>%
  group_by(shape) %>%
  anova_test(m_prop_look_dev ~ location_deviant)

anova_location %>% get_anova_table()

```
```{r}
anova_full <- aux_prop_look_dev_condition %>%
  filter(!participantid == 136) %>%
  anova_test(m_prop_look_dev ~ shape * deviant_type * orientation * location_deviant)

get_anova_table(anova_full)

```


# Save data

```{r}
setwd(path_data_processed)
write.csv(aux_prop_look_dev, file = "data_output_step05_individual_summary.csv", row.names = FALSE)
```

```{r}
setwd(path_data_processed)
write.csv(aux_prop_look_dev_group, file = "data_output_step05_group_summary.csv", row.names = FALSE)
```


 

